An open-source 1D electromagnetic modeler in Python: empymod

### 2017-04-03 ManuscriptGeophysics2016.tex
Re-submission version

\title{An open-source \new{1D }electromagnetic modeler in Python: empymod}

@@

\footer{}
\lefthead{Werthmüller}
\righthead{Open-source \new{1D} EM modeler in Python}

@@

%%fakesection ===    ABSTRACT    ===
\begin{abstract} % 1-2 sentence(s) each
\old{
  The free software empymod combines two earlier presented algorithms in this
  journal, creating a new code written in Python that is faster and leaner.
  The main objective is to present a three-dimensional layered-earth
  electromagnetic modeler with vertical transverse isotropy in an easy
  accessible programming language, both in terms of costs and learning curve,
  and in a collaborative way. The code is hosted in a web-based Git repository
  under a lax permissive license, allowing anyone to use it, even for
  commercial purposes, as well as to contribute to its development.
  Comparisons show that this code is as precise and faster than the codes it is
  based upon, thanks to the use of different Hankel transforms and the
  throughout calculation.  This code might certainly be useful for
  professionals in the electromagnetic area, but I specifically hope this code
  to be useful for educational purposes.
} % Note: replace:
  % ``vectorization of the calculation'' with ``calculation as arrays''.
%
\new{%
  The presented  Python-code empymod computes the three-dimensional
  electromagnetic field in a layered-earth with vertical transverse isotropy
  by combining and extending two earlier presented algorithms in this journal.
  The bottleneck in frequency- and time-domain calculations of electromagnetic
  responses derived in the wavenumber-frequency domain is the transformations
  from wavenumber to space domain and from frequency to time domain, the
  so-called Hankel and Fourier transforms.  Three different Hankel transform
  methods (quadrature, quadrature-with-extrapolation, and filters) and four
  different Fourier transform methods (FFT, FFTLog,
  quadrature-with-extrapolation, and filters) are included in empymod, which
  allows to compare these different methods in terms of speed and precision.
  The best transform in terms of speed and precision depends on the modeled
  frequencies: available filters, for instance, are very fast and precise for
  frequencies in the range of controlled-source electromagnetic data, but fail
  in the frequency range of ground penetrating radar. Conventional quadrature,
  on the other hand, is in comparison very slow but can model any frequency.
  Examples comparing empymod with analytical solutions and with existing
  electromagnetic modelers illustrate the capabilities of empymod.
}
\end{abstract}

@@

% 1.1 CSEM in hydrocarbon exploration
The potential of electromagnetic methods for the detection of hydrocarbon
reservoirs is known for some decades, see for instance \cite{PIEEE.89.Nekut} or
\cite{B.SEG.91.Chave}. More recent, good overviews of the methodology and its
applications are give by \cite{SG.05.Edwards} and \cite{IEEE.12.Ziolkowksi}.
Whereas in the early days everything happened in idealized, isotropic one
dimensional (1D) earth models, it is generally agreed that the often complex
geology of hydrocarbon reservoirs requires two dimensional (2D) and three
dimensional (3D), anisotropic forward modelers. However, 1D models are still
very important\new{.} \old{, not last because they are very fast. Many problems
can be simplified to and henceforth solved with 1D models. More importantly, 1D
allows} \new{Besides that their calculation is very fast they allow} to study
single, isolated effects to the electromagnetic field, which is a crucial
foundation in understanding the electromagnetic field behaviour and a necessity
for understanding the phenomena at higher dimensions. 1D models are furthermore
often used in inversion routines of higher dimensions, for instance to generate
a starting model or \new{to calculate the primary fields for 2D/3D modeling and
inversion routines.} \old{embed a 3D body in a 1D background.}

% 1.2 Existing 1D solutions
Solutions for the electromagnetic fields in a layered-earth model have been
solved and published extensively using different approaches. The importance and
wide\-spread use of 1D models is shown in the continuous stream of publications
in this area, even in recent years. \cite{GJI.07.Loseth} solved the problem
with the scattering matrix formulation for a 1D earth with general anisotropy,
spanning the frequency range from controlled-source electromagnetics (CSEM) to
ground-pe\-ne\-tra\-ting radar (GPR).  \cite{GJI.09.Chave} presented a solution
in terms of independent and unique transverse electric (TE) and tangential
magnetic (TM) modes for the electrical isotropic case using the diffusive
approximation (without displacement currents, valid for low frequencies such as
in CSEM). \cite{GEO.09.Key} demonstrated why 1D models still matter by testing,
for instance, the benefit of additional frequencies in an inversion
routine\old{. Key}\new{; he} follows and extends the magnetic vector potential
approach by \cite{B.AP.82.Wait} for forward modeling, using the isotropic,
diffusive low-frequency approach.  \cite{GEO.15.Hunziker} obtained the
electromagnetic field in a layered earth with vertical transverse iso\-tro\-py
(VTI) by solving two equivalent scalar equations with a scalar global
reflection coefficient.  All of the four citations regarding 1D solutions have
quite extensive reference lists about the history of 1D forward modeling, the
last one featuring an interesting review of the history of 1D electromagnetic
derivations spanning almost 200 years.

% 1.3 Hankel transform
A crucial as well as very interesting part of electromagnetic modeling is,
once one moves from the theoretical derivation to the numerical implementation,
the Hankel transform involved in the transformation from the
wave\-num\-ber-fre\-que\-ncy domain to the space-frequency domain. 
\new{Because what is common in all these approaches is that they derive
solutions in the wavenumber-frequency domain. This in turn requires a Hankel
transform, which is a numerically expensive, infinite integral containing
oscillating, slowly decaying Bessel functions (the Hankel transform is also
known as the Fourier-Bessel transform).} The use of digital filters is quite
common in geophysics, known as the fast Hankel transform method (FHT), as
introduced by \cite{GP.71.Gosh}, and popularized by \cite{TRP.75.Anderson,
GEO.79.Anderson, TMS.82.Anderson} thanks to his freely available Fortran
routines. Naturally, standard quadrature can be used as well for the Hankel
transform, see for instance \cite{GEO.83.Chave}, and hybrid routines using both
methods were published by \cite{GEO.84.Anderson, GEO.89.Anderson}. The topic
got picked up again recently with new filters being published by
\cite{GP.07.Kong} and \cite{GEO.09.Key, GEO.12.Key}

@@

% 1.5 Claim
With empymod I present a 1D forward modeling code that is based on
\cite{GEO.15.Hunziker} for the wavenumber-frequency domain calculation, and
\new{mainly} on \cite{GEO.12.Key} for the Hankel and Fourier transforms.
\old{As I will therefore refer to these publications quite a lot, I use the
name Hun15 to denote Hunziker et al. (2015), and the names Key09 and Key12 to
denote Key (2009, 2012). To denote the FHT filters as published by Key09 and
Key12 I will add the corresponding filter-size, e.g. Key09-401.}
In addition \new{to QWE and FHT}, empymod includes \new{an
adaptive quadrature (QUAD) for the Hankel transform, and the standard fast
Fourier transform FFT as well as} the logarithmic fast Fourier transform
FFTLog of \cite{RAS.00.Hamilton} \old{as a third} \new{for the} Fourier
transform \old{possibility to FHT and QWE for the frequency-to-time
transformation}.

To my knowledge, empymod is the first code that is freely available which
calculates the full wavefield for a layered-earth model with vertical isotropy
and has \old{both quadrature and filters} \new{different types of Hankel
transforms (QWE, QUAD, FHT) and various methods for the Fourier transform
(FFTLog, sine/cosine-filters, QWE, FFT)}\old{built in}. \old{This makes}
\new{An interesting use-case for} empymod \old{ideal not just for} \new{is
therefore} comparison studies of the \old{two} \new{different} methods,
\old{but also to build} \new{up to building} hybrid inversion schemes.
\old{Further advantages of empymod are:} \new{The code is published under the
lax permissive \emph{Apache Version 2.0 license}, which makes it available to
everyone\new{ for free}, even for commercial purposes. It therefore might
prove to be valuable for students and professionals alike. It is written in
Python, a modern, cross-platform, free and open-source programming language.
With its scientific libraries, mainly the numeric and scientific modules
\texttt{NumPy} and \texttt{SciPy}, it creates an extremely powerful numerical
calculation stack: Even though it is an interpreted language it can be very
fast, as \texttt{NumPy}/\texttt{SciPy} use under the hood routines in Fortran
or in C (using for instance the BLAS/Lapack libraries) for the linear algebra
computations, and Python is merely the glue. The comparisons show that
empymod is as fast as the existing codes. The code is vectorized (avoiding
loops) as much as possible, which improves computation. The most time-consuming
calculations have furthermore a flag to run parallelized. However, on a larger
scale such as an inversion routine one would probably want to parallelize the
kernel calls instead of the kernel itself. The code is hosted on GitHub, which
makes it easy for anyone to improve and contribute to the code:
\url{https://github.com/prisae/empymod}. The documentation is hosted on
\url{https://empymod.readthedocs.io}.}
\cite{TLE.17.Werthmuller}\new{ gives a tutorial-style introduction to EM modeling
using empymod.}

\old{
\textbf{(1) Python:} Python is a modern, cross-platform, free and open-source
programming language. With its scientific libraries, mainly the numeric and
scientific modules \texttt{NumPy} and \texttt{SciPy}, it creates an extremely
powerful numerical calculation stack.
}

\old{
\textbf{(2) Libre (free and open-source):} The code is published under the lax
permissive \emph{Apache Version 2.0 license}, which makes it available to
everyone, even for commercial purposes. It therefore might prove to be valuable
for students and professionals alike. Not only the code is free, but also the
underlying platform (Python), contrary to, for instance, Matlab. As such it
is ideal for reproducible research.
}

\old{
\textbf{(3) Lean:} The codebase is very lean. It is built from scratch, and it
therefore does not suffer the problems that sometimes come with the organic
growth of a codebase. The code follows as much as possible the \emph{DRY}
coding paradigm, \emph{Don't Repeat Yourself}.
}

\old{
\textbf{(4) Fast:} The code is vectorized \new{(avoiding loops)} as much as
possible, which improves computation. The most time-consuming calculations have
furthermore a flag to run parallelized. However, on a larger scale such as an
inversion routine one would probably want to parallelize the kernel calls
instead of the kernel itself. Even though Python is an interpreted language it
can be very fast, as \new{under the hood routines} \old{the underlying routines
are} in Fortran or in C (using for instance the BLAS/Lapack libraries)
\new{are used for the linear algebra computations}, and Python is merely the
glue. The comparisons show that empymod is as fast as the existing codes.
}

\old{
\textbf{(5) Community:} The code is hosted on GitHub, which makes it easy for
anyone to improve and contribute to the code, and will allow empymod to
grow:\\
\url{https://github.com/prisae/empymod}.
}

\old{
\textbf{(6) Well documented:} The code is extensively documented, and large
parts of it are extracted using automated tools (Sphinx) to create a manual:\\
\url{https://empymod.readthedocs.io}.
}

\old{These points make this a worthwhile extension to the existing codes from
Hun15 and Key12. The existing codes are written in Fortran, which requires
much more time than Python to get started for students or to develop for
anyone, or Matlab, which is a proprietary language. And the existing codes are
in static repositories where one can only download the code, which makes
interaction, contribution, or bug filing more difficult.}

After introducing the code in the first part I will present some comparisons by
reproducing results from the publications this code is based upon: First
\old{a} \new{some} comparison\new{s} to the analytical half-space solution,
followed by a comparison to EMmod by \old{Hun15}\cite{GEO.15.Hunziker}, and
finally a comparison to some results presented by \old{Key12}\cite{GEO.12.Key}.
\new{Last is a time-domain example in which I compare the four different
  Fourier transforms.}

@@

The code consists of 5 files; these are 3 core modules plus \emph{utils}, which
contains input checks and other utilities, and \emph{filters}, containing the
FHT filter coefficients. The three core routines are: (1) \emph{kernel}, where
the wavenumber-domain calculation is carried out; (2) \emph{transform}, where
the Hankel and Fourier transforms are computed; and (3) \emph{model}, which
contains the actual modeling routines for end users.
\new{The main modeling routine is \emph{bipole}, which can calculate frequency-
and time-domain responses for arbitrary oriented, electric or magnetic bipole
sources and receivers of finite length.}
\old{As of now there are two main modeling routines implemented,
\emph{frequency} and \emph{time}, with which one can calculate the frequency-
and time-domain responses for electric or magnetic point sources and receivers,
directed along the three principal axis $x, y,$ and $z$. More modeling routines
can easily be added to \emph{model}, such as finite bipoles or arbitrary source
and receiver directions, as they do not affect the core of the calculation,
hence not \emph{kernel} nor \emph{transform}.}

The wavenumber-domain calculation in \emph{kernel} follows
\old{Hun15}\cite{GEO.15.Hunziker}, and calculates as such the complete
wavefield for a layered VTI model\new{.} \old{,
where the} \new{The} code makes no assumptions about the model\old{.}\new{:}
Source and receiver can be placed anywhere in the model\old{.}\new{, including
first and last layer, and you have to define if the first layer is air or
not. Bipoles can cross layer boundaries.} Depths, frequencies, and
source-receiver configuration have to be defined, and each layer is
characterized with its horizontal resistivity $\rho_\textrm{h}$, its electrical
anisotropy $\lambda$, where $\lambda = \sqrt{\rho_\textrm{v}/\rho_\textrm{h}}$,
its horizontal and vertical magnetic permeabilities $\mu_\mathrm{h}$ and
$\mu_\mathrm{v}$, and the horizontal and vertical electric permittivities
$\epsilon_\mathrm{h}$ and $\epsilon_\mathrm{v}$. The main difference\old{s}
between EMmod and empymod \new{is}\old{are the programming language and} the
Hankel transform, and a few additional fundamental differences: EMmod uses 2nd
order Bessel functions of the first kind $J_2$.  Published FHT filters
generally provide coefficients for 0th and 1st order Bessel functions $J_0,
J_1$ only.  Therefore, the recurrence relation
%
\begin{equation}
  J_2(kr) = \frac{2}{kr}J_1(kr) - J_0(kr)
  \label{eq:j2}
\end{equation}
%
is used, where $k$ and $r$ are the wavenumber and space-domain parameters,
respectively.
\new{Other differences are that empymod includes Fourier transforms, hence
time-domain calculation, and can model arbitrary rotated, finite bipoles.}
\old{Another difference is vectorization: EMmod carries out the
calculation in the wavenumber-domain by looping over frequencies, offsets, and
wavenumbers. This is carried out in a single calculation without looping in
empymod. Another difference is that empymod is much leaner than EMmod.
EMmod grew organically, as confirmed by the author, adding more and more
features, where empymod was designed with all 36 source-receiver
configurations from the beginning. As an example, the full-space solution in
empymod is one function (107 lines of code) for all source-receiver
configurations, whereas in EMmod it is split into no less than 16 functions
(364 lines of code). This should help to maintain the code easier, and it
should also be easier for new contributors to read into the code.}

The \new{QWE and filter} Hankel and Fourier transforms in \emph{transform}
follow \old{Key12}\cite{GEO.12.Key}, with a few changes. The most important
ones regarding speed is vectorization\new{, hence avoiding loops,} and a
splined version for the filter method, in addition to the traditional lagged
version: The lagged version samples the wavenumber from the minimum to the
maximum required wavenumber given the required offsets and the chosen filter
base with the spacing as defined by the filter. It subsequently carries out the
Hankel transform, and interpolates for the required offsets in the space
domain. The splined version, on the other hand, uses a user-specified number of
values per decade from the minimum to the maximum wavenumber. It then
interpolates for the required wavenumber values, and does the Hankel transform
afterwards. The lagged version is \emph{very} fast. However, with the splined
version a speed-up can be achieved in comparison with the original FHT at
higher precision if compared with the lagged version. All filters published
\new{in the source codes of }\old{by} \old{Key09 and Key12}\cite{GEO.09.Key,
GEO.12.Key} are included in empymod, which includes Key's own filters as well
as the filters by \cite{TMS.82.Anderson} and by \cite{GP.07.Kong}.  \new{(I
refer throughout to paper to a specific filter with author, year, and
filter-size, for instance Key09-201.)}

The most import part of \old{Key12}\new{GEO.12.Key} is the introduction of a
new quadrature algorithm to geophysics, named
qua\-dra\-ture-with-ex\-tra\-po\-la\-tion (QWE). QWE is a fast quadrature
method using the Shanks transformation \citep{JMP.55.Shanks} computed with
Wynn's epsilon algorithm \citep{MC.56.Wynn}. The advantage of quadrature over
filters is the ability to estimate the error. QWE continues until the absolute
error, estimated by the difference of subsequent iterations $n$, satisfies the
inequality
%
\begin{equation}
  |S^*_n-S^*_{n-1}| \le \varepsilon_r|S^*_n| + \varepsilon_a\ ,
  \label{eq:err}
\end{equation}
%
where $S^*$ is the extrapolated result, and $\varepsilon_r, \varepsilon_a$ are
the relative and absolute tolerance, respectively.

Whereas for the Hankel transform the \old{two}\new{three} methods QWE\new{,}
\old{and} FHT\new{, and QUAD, a standard adaptive quadrature using QAGSE from
the Fortran QUADPACK library,} are implemented, for the Fourier transform there
are \old{three}\new{four} methods implemented: QWE, FHT with the sine and
cosine filters, \new{the standard FFT,} and FFTLog \citep{RAS.00.Hamilton}.
The logarithmic fast Fourier transform FFTLog is ideal for this operation, as
generally a wide range of frequencies is required to go from \new{the}
frequency to \new{the} time domain. The FFTLog can yield faster results than
the QWE method and more precise than the FHT results, specifically for land
impulse responses at early times.

\old{In addition to} \new{Calculation time can be reduced by using} the splined
version of the QWE and the splined and lagged versions of the FHT\old{,
a}\new{. In addition, a}ll time-consuming calculations are set up to be able to
run in parallel, with the help of the Python-module numexpr.  This can
significantly speed up calculations if you run big models with many layers and
for many offsets and frequencies. However, if you include empymod in an
inversion scheme then it might be better to parallelize the calls to empymod,
instead of empymod itself.

It is important to note that calculations in the wavenumber domain depend
\emph{only} on offset $r = \sqrt{x^2+y^2}$; the scaling factor, which depends
on the angle $\varphi = \rm{atan2}(y, x)$, is multiplied afterwards. In order
to calculate for instance a circle around the source, the kernel has to be
called only once, and subsequently scaled by the angle-dependent factor for
each source-receiver pair. This makes the splined version very powerful, as it
can be used for irregularly distributed data\new{ as long as all receiver
depths are the same}.

\old{The code distributed on GitHub contains Jupyter Notebooks to reproduce the
results in this article, as well as some basic tests and benchmarks, and the
\LaTeX-source of the article itself.}

The run-time comparisons tests were run on a Lenovo ThinkCentre running Ubuntu
16.04 64-bit, with 8\,GB of memory and an Intel Core i7-4770 CPU @ 3.40GHz x
8\old{ (4 cores with hyper-threading)}. Comparing run times is always a
difficult task, and between different languages even more, here between
Python \new{(empymod)},
Fortran \new{(EMmod, Dipole1D)},
and Matlab \new{(scripts from \cite{GEO.12.Key})}.
However, they do serve as comparison.  I used the
Matlab-\texttt{timing} and Python-\texttt{timeit} functions, which behave
very similar. It is probably expected today, but still worth mentioning, that
the calculation is carried out in double precision.

@@

\cite{PIER.10.Slob} published analytical frequency- and time-domain solutions
for the diffusive electric field in a VTI half-space\new{, where source and
receiver can be placed anywhere in the half-space}. As an example, I use the
same model as \old{Hun15}\cite{GEO.15.Hunziker} in their Figures~1 and~2: A
half-space with horizontal resistivity $\rho_\mathrm{h} = 1/3\,\Omega$\,m,
anisotropy $\lambda = \sqrt{10}$, and for frequency $f = 0.5\,$Hz. The source
is located horizontally at the origin \new{($x_\mathrm{s} = y_\mathrm{s} =
0\,$m)} at a depth of 150\,m, and the depth of the receivers is 200\,m. The
field is calculated on a regular grid with a spacing of 10\,m.
Figure~\ref{fig:analytical} shows the analytical amplitude and phase results
for the first quadrant (the other quadrants are simply symmetric copies of it).
%
\plot*[btp]{analytical}{width=\textwidth}{Analytical solution for a
half-space with $\rho_\mathrm{h} = 1/3\,\Omega$\,m, $\lambda = \sqrt{10}$, $f =
0.5\,$Hz, $z_\mathrm{s} = 150\,$m, and $z_\mathrm{r} = 200\,$m, calculated on a
regular grid with a spacing of 10 meters; (a) amplitude and (b) phase.}
%
The figure shows the result for x-directed electric source and receiver
dipoles ($G^\mathrm{ee}_{xx}$), other configurations yield similar results (the
choice is based on the insight that this is the most interesting of all
electric source to electric receiver fields, as can be seen in
\old{Hun15}\cite{GEO.15.Hunziker} Figures~1 and~2).

>>>>

\title{An open-source 1D electromagnetic modeler in Python: empymod}

@@

\footer{}
\lefthead{Werthmüller}
\righthead{Open-source 1D EM modeler in Python}

@@

%%fakesection ===    ABSTRACT    ===
\begin{abstract} % 1-2 sentence(s) each
  The presented Python-code empymod computes the three-dimensional
  electromagnetic field in a layered-earth with vertical transverse isotropy by
  combining and extending two earlier presented algorithms in this journal.
  The bottleneck in frequency- and time-domain calculations of electromagnetic
  responses derived in the wavenumber-frequency domain is the transformations
  from wavenumber to space domain and from frequency to time domain, the
  so-called Hankel and Fourier transforms. Three different Hankel transform
  methods (quadrature, quadrature-with-extrapolation, and filters) and four
  different Fourier transform methods (FFT, FFTLog,
  quadrature-with-extrapolation, and filters) are included in empymod, which
  allows to compare these different methods in terms of speed and precision.
  The best transform in terms of speed and precision depends on the modeled
  frequencies: available filters, for instance, are very fast and precise for
  frequencies in the range of controlled-source electromagnetic data, but fail
  in the frequency range of ground penetrating radar. Conventional quadrature,
  on the other hand, is in comparison very slow but can model any frequency.
  Examples comparing empymod with analytical solutions and with existing
  electromagnetic modelers illustrate the capabilities of empymod.
\end{abstract}

@@

% 1.1 CSEM in hydrocarbon exploration
The potential of electromagnetic methods for the detection of hydrocarbon
reservoirs is known for some decades, see for instance \cite{PIEEE.89.Nekut} or
\cite{B.SEG.91.Chave}. More recent, good overviews of the methodology and its
applications are give by \cite{SG.05.Edwards} and \cite{IEEE.12.Ziolkowksi}.
Whereas in the early days everything happened in idealized, isotropic one
dimensional (1D) earth models, it is generally agreed that the often complex
geology of hydrocarbon reservoirs requires two dimensional (2D) and three
dimensional (3D), anisotropic forward modelers. However, 1D models are still
very important. Besides that their calculation is very fast they allow to study
single, isolated effects to the electromagnetic field, which is a crucial
foundation in understanding the electromagnetic field behaviour and a necessity
for understanding the phenomena at higher dimensions. 1D models are furthermore
often used in inversion routines of higher dimensions, for instance to generate
a starting model or to calculate the primary fields for 2D/3D modeling and
inversion routines.

% 1.2 Existing 1D solutions
Solutions for the electromagnetic fields in a layered-earth model have been
solved and published extensively using different approaches. The importance and
wide\-spread use of 1D models is shown in the continuous stream of publications
in this area, even in recent years. \cite{GJI.07.Loseth} solved the problem
with the scattering matrix formulation for a 1D earth with general anisotropy,
spanning the frequency range from controlled-source electromagnetics (CSEM) to
ground-pe\-ne\-tra\-ting radar (GPR). \cite{GJI.09.Chave} presented a solution
in terms of independent and unique transverse electric (TE) and tangential
magnetic (TM) modes for the electrical isotropic case using the diffusive
approximation (without displacement currents, valid for low frequencies such as
in CSEM). \cite{GEO.09.Key} demonstrated why 1D models still matter by testing,
for instance, the benefit of additional frequencies in an inversion routine; he
follows and extends the magnetic vector potential approach by
\cite{B.AP.82.Wait} for forward modeling, using the isotropic, diffusive
low-frequency approach. \cite{GEO.15.Hunziker} obtained the electromagnetic
field in a layered earth with vertical transverse iso\-tro\-py (VTI) by solving
two equivalent scalar equations with a scalar global reflection coefficient.
All of the four citations regarding 1D solutions have quite extensive reference
lists about the history of 1D forward modeling, the last one featuring an
interesting review of the history of 1D electromagnetic derivations spanning
almost 200 years.

% 1.3 Hankel transform
A crucial as well as very interesting part of electromagnetic modeling is, once
one moves from the theoretical derivation to the numerical implementation, the
Hankel transform involved in the transformation from the
wave\-num\-ber-fre\-que\-ncy domain to the space-frequency domain. Because
what is common in all these approaches is that they derive solutions in the
wavenumber-frequency domain. This in turn requires a Hankel transform, which is
a numerically expensive, infinite integral containing oscillating, slowly
decaying Bessel functions (the Hankel transform is also known as the
Fourier-Bessel transform). The use of digital filters is quite common in
geophysics, known as the fast Hankel transform method (FHT), as introduced by
\cite{GP.71.Gosh}, and popularized by \cite{TRP.75.Anderson, GEO.79.Anderson,
TMS.82.Anderson} thanks to his freely available Fortran routines. Naturally,
standard quadrature can be used as well for the Hankel transform, see for
instance \cite{GEO.83.Chave}, and hybrid routines using both methods were
published by \cite{GEO.84.Anderson, GEO.89.Anderson}. The topic got picked up
again recently with new filters being published by \cite{GP.07.Kong} and
\cite{GEO.09.Key, GEO.12.Key}

@@

% 1.5 Claim
With empymod I present a 1D forward modeling code that is based on
\cite{GEO.15.Hunziker} for the wavenumber-frequency domain calculation, and
mainly on \cite{GEO.12.Key} for the Hankel and Fourier transforms. In addition
to QWE and FHT, empymod includes an adaptive quadrature (QUAD) for the Hankel
transform, and the standard fast Fourier transform FFT as well as the
logarithmic fast Fourier transform FFTLog of \cite{RAS.00.Hamilton} for the
Fourier transform.

To my knowledge, empymod is the first code that is freely available which
calculates the full wavefield for a layered-earth model with vertical isotropy
and has different types of Hankel transforms (QWE, QUAD, FHT) and various
methods for the Fourier transform (FFTLog, sine/cosine-filters, QWE, FFT). An
interesting use-case for empymod is therefore comparison studies of the
different methods, up to building hybrid inversion schemes. The code is
published under the lax permissive \emph{Apache Version 2.0 license}, which
makes it available to everyone for free, even for commercial purposes. It
therefore might prove to be valuable for students and professionals alike. It
is written in Python, a modern, cross-platform, free and open-source
programming language. With its scientific libraries, mainly the numeric and
scientific modules \texttt{NumPy} and \texttt{SciPy}, it creates an extremely
powerful numerical calculation stack: Even though it is an interpreted language
it can be very fast, as \texttt{NumPy}/\texttt{SciPy} use under the hood
routines in Fortran or in C (using for instance the BLAS/Lapack libraries) for
the linear algebra computations, and Python is merely the glue. The comparisons
show that empymod is as fast as the existing codes. The code is vectorized
(avoiding loops) as much as possible, which improves computation. The most
time-consuming calculations have furthermore a flag to run parallelized.
However, on a larger scale such as an inversion routine one would probably want
to parallelize the kernel calls instead of the kernel itself. The code is
hosted on GitHub, which makes it easy for anyone to improve and contribute to
the code: \url{https://github.com/prisae/empymod}. The documentation is hosted
on \url{https://empymod.readthedocs.io}. \cite{TLE.17.Werthmuller} gives a
tutorial-style introduction to EM modeling using empymod.

After introducing the code in the first part I will present some comparisons by
reproducing results from the publications this code is based upon: First some
comparisons to the analytical half-space solution, followed by a comparison to
EMmod by \cite{GEO.15.Hunziker}, and finally a comparison to some results
presented by \cite{GEO.12.Key}. Last is a time-domain example in which I
compare the four different Fourier transforms.

@@

The code consists of 5 files; these are 3 core modules plus \emph{utils}, which
contains input checks and other utilities, and \emph{filters}, containing the
FHT filter coefficients. The three core routines are: (1) \emph{kernel}, where
the wavenumber-domain calculation is carried out; (2) \emph{transform}, where
the Hankel and Fourier transforms are computed; and (3) \emph{model}, which
contains the actual modeling routines for end users. The main modeling routine
is \emph{bipole}, which can calculate frequency- and time-domain responses for
arbitrary oriented, electric or magnetic bipole sources and receivers of finite
length.

The wavenumber-domain calculation in \emph{kernel} follows
\cite{GEO.15.Hunziker}, and calculates as such the complete wavefield for a
layered VTI model. The code makes no assumptions about the model: Source and
receiver can be placed anywhere in the model, including first and last layer,
and you have to define if the first layer is air or not. Bipoles can cross
layer boundaries. Depths, frequencies, and source-receiver configuration have
to be defined, and each layer is characterized with its horizontal resistivity
$\rho_\textrm{h}$, its electrical anisotropy $\lambda$, where $\lambda =
\sqrt{\rho_\textrm{v}/\rho_\textrm{h}}$, its horizontal and vertical magnetic
permeabilities $\mu_\mathrm{h}$ and $\mu_\mathrm{v}$, and the horizontal and
vertical electric permittivities $\epsilon_\mathrm{h}$ and
$\epsilon_\mathrm{v}$. The main difference between EMmod and empymod is the
Hankel transform, and a few additional fundamental differences: EMmod uses 2nd
order Bessel functions of the first kind $J_2$. Published FHT filters
generally provide coefficients for 0th and 1st order Bessel functions $J_0,
J_1$ only. Therefore, the recurrence relation
%
\begin{equation}
  J_2(kr) = \frac{2}{kr}J_1(kr) - J_0(kr)
  \label{eq:j2}
\end{equation}
%
is used, where $k$ and $r$ are the wavenumber and space-domain parameters,
respectively. Other differences are that empymod includes Fourier transforms,
hence time-domain calculation, and can model arbitrary rotated, finite bipoles.

The QWE and filter Hankel and Fourier transforms in \emph{transform} follow
\cite{GEO.12.Key}, with a few changes. The most important ones regarding speed
is vectorization, hence avoiding loops, and a splined version for the filter
method, in addition to the traditional lagged version: The lagged version
samples the wavenumber from the minimum to the maximum required wavenumber
given the required offsets and the chosen filter base with the spacing as
defined by the filter. It subsequently carries out the Hankel transform, and
interpolates for the required offsets in the space domain. The splined version,
on the other hand, uses a user-specified number of values per decade from the
minimum to the maximum wavenumber. It then interpolates for the required
wavenumber values, and does the Hankel transform afterwards. The lagged version
is \emph{very} fast. However, with the splined version a speed-up can be
achieved in comparison with the original FHT at higher precision if compared
with the lagged version. All filters published in the source codes of
\cite{GEO.09.Key, GEO.12.Key} are included in empymod, which includes Key's own
filters as well as the filters by \cite{TMS.82.Anderson} and by
\cite{GP.07.Kong}. (I refer throughout to paper to a specific filter with
author, year, and filter-size, for instance Key09-201.)

The most import part of \cite{GEO.12.Key} is the introduction of a new
quadrature algorithm to geophysics, named
qua\-dra\-ture-with-ex\-tra\-po\-la\-tion (QWE). QWE is a fast quadrature
method using the Shanks transformation \citep{JMP.55.Shanks} computed with
Wynn's epsilon algorithm \citep{MC.56.Wynn}. The advantage of quadrature over
filters is the ability to estimate the error. QWE continues until the absolute
error, estimated by the difference of subsequent iterations $n$, satisfies the
inequality
%
\begin{equation}
  |S^*_n-S^*_{n-1}| \le \varepsilon_r|S^*_n| + \varepsilon_a\ ,
  \label{eq:err}
\end{equation}
%
where $S^*$ is the extrapolated result, and $\varepsilon_r, \varepsilon_a$ are
the relative and absolute tolerance, respectively.

Whereas for the Hankel transform the three methods QWE, FHT, and QUAD, a
standard adaptive quadrature using QAGSE from the Fortran QUADPACK library, are
implemented, for the Fourier transform there are four methods implemented: QWE,
FHT with the sine and cosine filters, the standard FFT, and FFTLog
\citep{RAS.00.Hamilton}. The logarithmic fast Fourier transform FFTLog is
ideal for this operation, as generally a wide range of frequencies is required
to go from the frequency to the time domain. The FFTLog can yield faster
results than the QWE method and more precise than the FHT results, specifically
for land impulse responses at early times.

Calculation time can be reduced by using the splined version of the QWE and the
splined and lagged versions of the FHT. In addition, all time-consuming
calculations are set up to be able to run in parallel, with the help of the
Python-module numexpr. This can significantly speed up calculations if you run
big models with many layers and for many offsets and frequencies. However, if
you include empymod in an inversion scheme then it might be better to
parallelize the calls to empymod, instead of empymod itself.

It is important to note that calculations in the wavenumber domain depend
\emph{only} on offset $r = \sqrt{x^2+y^2}$; the scaling factor, which depends
on the angle $\varphi = \rm{atan2}(y, x)$, is multiplied afterwards. In order
to calculate for instance a circle around the source, the kernel has to be
called only once, and subsequently scaled by the angle-dependent factor for
each source-receiver pair. This makes the splined version very powerful, as it
can be used for irregularly distributed data as long as all receiver depths are
the same.

The run-time comparisons tests were run on a Lenovo ThinkCentre running Ubuntu
16.04 64-bit, with 8\,GB of memory and an Intel Core i7-4770 CPU @ 3.40GHz x 8.
Comparing run times is always a difficult task, and between different languages
even more, here between Python (empymod), Fortran (EMmod, Dipole1D), and Matlab
(scripts from \cite{GEO.12.Key}). However, they do serve as comparison. I
used the Matlab-\texttt{timing} and Python-\texttt{timeit} functions, which
behave very similar. It is probably expected today, but still worth mentioning,
that the calculation is carried out in double precision.

@@

\cite{PIER.10.Slob} published analytical frequency- and time-domain solutions
for the diffusive electric field in a VTI half-space, where source and receiver
can be placed anywhere in the half-space. As an example, I use the same model
as \cite{GEO.15.Hunziker} in their Figures~1 and~2: A half-space with
horizontal resistivity $\rho_\mathrm{h} = 1/3\,\Omega$\,m, anisotropy $\lambda
= \sqrt{10}$, and for frequency $f = 0.5\,$Hz. The source is located
horizontally at the origin ($x_\mathrm{s} = y_\mathrm{s} = 0\,$m) at a depth of
150\,m, and the depth of the receivers is 200\,m. The field is calculated on a
regular grid with a spacing of 10\,m. Figure~\ref{fig:analytical} shows the
analytical amplitude and phase results for the first quadrant (the other
quadrants are simply symmetric copies of it).
%
\plot*[btp]{analytical}{width=\textwidth}{Analytical solution for a half-space
  with $\rho_\mathrm{h} = 1/3\,\Omega$\,m, $\lambda = \sqrt{10}$, $f =
  0.5\,$Hz, $z_\mathrm{s} = 150\,$m, and $z_\mathrm{r} = 200\,$m, calculated on
  a regular grid with a spacing of 10 meters; (a) amplitude and (b) phase.}
%
The figure shows the result for x-directed electric source and receiver dipoles
($G^\mathrm{ee}_{xx}$), other configurations yield similar results (the choice
is based on the insight that this is the most interesting of all electric
source to electric receiver fields, as can be seen in \cite{GEO.15.Hunziker}
Figures~1 and~2).
